Translation of B implementations to the LLVM (\textsc{draft})

### 2012-09-04 b0-llvm.tex
Adding some fix until 5.3 Relations

We first define the translation rule for the root of the source
program abstract syntax tree, which shall correspond to a B
implementation. This is a preliminary version where the code
generation is only able to handle standalone B0 implementations, i.e.
the state is defined in the ``concrete variables'' clause. Moreover
the data is restricted to predefined basic types \textsf{INT} and
\textsf{BOOL}:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{impl} : & \textsf{ Node} \rightarrow text \\
  \trad{i}{impl} \isdef & \textbf{ let } O_{c} = \trad{n.values}{val} \textbf{ and } O_v = \trad{n.vars}{var} \\
  & \quad \textbf{ and } O_{in} = \trad{n.initialization}{ini} \textbf{ and } O_{op} = \trad{n.ops}{ops} \\
  & \quad \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad \trad{i.ops}{ops}  \\
  & \textbf{ where } \textsf{i.k} = \textsf{Impl} 
\end{split}
\end{align}
\end{framed}
Code generation for initialisation and operations is specified in
section~\ref{sec:control}. Code generation for data (type and
state) is defined in section~\ref{sec:data}, and for expression
evaluation in section~\ref{sec:expr}.

@@

Concrete variables are mapped to so-called internal variables in the
LLVM program, i.e. private data that is local to the produced
file. The following rule describes the code generation for such
variables:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{var} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{type} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = internal global "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{type} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

Constant valuations are translated to constants in the
LLVM program, as described in the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{val} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{val} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = constant "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{val} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

Based on this definition, the translation of boolean literals is straightforward:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D_{bool}(\textsf{n.val}), \texttt{"i1"} \\
  & \textbf{ where } \textsf{n.k} = \textsf{LitBool}
\end{split}
\end{align}
\end{framed}

The translation of integer literals is even simpler, as they have the
same denotation in B0 and LLVM:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, \textsf{n.val}, \texttt{"i32"} \\
  & \textbf{ where } \textsf{n.k} = \textsf{LitInt}
\end{split}
\end{align}
\end{framed}

@@

\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D(\textsf{n}), \trad{n.type}{type} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Ident}
\end{split}
\end{align}
\end{framed}

@@

Based upon this definition, the following rule specifies the
generation of LLVM code responsible for the evaluation of binary
arithmetic expressions (plus the name of the local variable storing
the resulting value, and of its LLVM type):
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
  & \quad \textbf{ and } v = \mathcal{V} \textbf{ and } op = D_{bari}(\textsf{n.op}) \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad v \conc \texttt{" = i32 "} \conc O_l \conc^, O'_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} \in \dom D_{bari} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

Next rule specifies the code generation for the successor and
predecessor operators. As LLVM has no corresponding statement, the
generated code uses addition and subtraction:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 add 1, "} \conc O_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{succ} \\
  & \textbf{ assert } O_t = \texttt{i32} \\
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 sub "} \conc O_l \conc \texttt{", 1"} \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{pred} \\
  & \textbf{ assert } O_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

The following rule specifies code generation for the unary
minus. Since LLVM does not have a specific statement for such
operation, the translation uses subtraction:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 sub 0, "} \conc O_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{-} \\
  & \textbf{ assert } O_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

Also, LLVM does not have a predefined operation for the ``power''
operator provided by B0. The corresponding rule introduces a call
to a function, named \texttt{@power} that implements this operator (the LLVM source code
for such function is not presented herein):
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
  & \quad \textbf{ and } l = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad l \conc \texttt{" = call i32 @power(i32, "} \conc O_l \conc \texttt{", i32 "} \conc O'_l \conc \texttt{")"}) \conc \nl), \\
  & \quad l, \texttt{"i32"} \\
  & \textbf{ where } \textsf{r.k} = \textsf{Power} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

@@

\begin{framed}
\begin{align}
\begin{split}
  \trad{r, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{r.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{r.right, $D$}{exp} \\
  & \quad \textbf{ and } l = \mathcal{V} \textbf{ and } op = D_{op}(\textsf{r.op}) \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad l \conc \texttt{" = icmp "} \conc op \conc \texttt{" "} \conc O_t \conc \texttt{" "} \conc O_l \conc^, O'_l \conc \nl), \\
  & \quad l \\
  & \textbf{ where } \textsf{r.k} = \textsf{BinRel} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

>>>>

We first define the translation rule for the root of the source
program abstract syntax tree, which shall correspond to a B
implementation. This is a preliminary version where the code
generation is only able to handle standalone B0 implementations, i.e.
the state is defined in the ``concrete variables'' clause. Moreover
the data is restricted to predefined basic types \textsf{INT} and
\textsf{BOOL}:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{impl} : & \textsf{ Node} \rightarrow text \\
  \trad{i}{impl} \isdef & \textbf{ let } O_{c} = \trad{i.values}{val} \textbf{ and } O_v = \trad{i.vars}{var} \\
  & \quad \textbf{ and } O_{in} = \trad{i.initialization}{ini} \textbf{ and } O_{op} = \trad{i.ops}{ops} \\
  & \quad \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad \trad{i.ops}{ops}  \\
  & \textbf{ where } \textsf{i.k} = \textsf{Impl} 
\end{split}
\end{align}
\end{framed}
Code generation for initialisation and operations is specified in
section~\ref{sec:control}. Code generation for data (type and
state) is defined in section~\ref{sec:data}, and for expression
evaluation in section~\ref{sec:expr}.

@@

Concrete variables are mapped to so-called internal variables in the
LLVM program, i.e. private data that is local to the produced
file. The following rule describes the code generation for such
variables:

% TODO: Qual o valor de n ?  Seria isto: n = L()  (\name) ?
% Nas duas funções seguintes parece ter uma mistura com
% relação ao nome das funções:  {var}, {type}, {val} 
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{var} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{type} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident}  \textbf{ and }  n = ???? \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = internal global "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{type} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

Constant valuations are translated to constants in the
LLVM program, as described in the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{val} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{val} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ and }  n = ???? \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = constant "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{val} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

Based on this definition, the translation of boolean literals is straightforward:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{exp} : & \textsf{ Node} \times dict \rightarrow text \times text \times  text  \\
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D_{bool}(\textsf{n.val}), \texttt{"i1"} \\
  & \textbf{ where } \textsf{n.k} = \textsf{LitBool}
\end{split}
\end{align}
\end{framed}

The translation of integer literals is even simpler, as they have the
same denotation in B0 and LLVM:
\begin{framed}
\begin{align}
\begin{split}
 \trad{}{exp} : & \textsf{ Node} \times dict \rightarrow text  \times text \times  text\\
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, \textsf{n.val}, \texttt{"i32"} \\
  & \textbf{ where } \textsf{n.k} = \textsf{LitInt}
\end{split}
\end{align}
\end{framed}

@@

\begin{framed}
\begin{align}
\begin{split}
  \trad{}{exp} : & \textsf{ Node} \times dict \rightarrow text \times text \times  text\\
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D(\textsf{n}), \trad{n.type}{type} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Ident}
\end{split}
\end{align}
\end{framed}

@@

Based upon this definition, the following rule specifies the
generation of LLVM code responsible for the evaluation of binary
arithmetic expressions (plus the name of the local variable storing
the resulting value, and of its LLVM type):
% TODO: Explicar o que significa o O_p,  O_l,   O_t  ?
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{exp} : & \textsf{ Node} \times dict  \rightarrow text \times text \times  text\\
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
  & \quad \textbf{ and } v = \name \textbf{ and } op = D_{bari}(\textsf{n.op}) \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad v \conc \texttt{" = i32 "} \conc O_l \conc^, O'_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} \in \dom D_{bari} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

Next rule specifies the code generation for the successor and
predecessor operators. As LLVM has no corresponding statement, the
generated code uses addition and subtraction:
\begin{framed}
\begin{align}
\begin{split}
 \trad{}{exp} : & \textsf{ Node} \times dict  \rightarrow text \times text \times  text\\
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \name \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 add 1, "} \conc O_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{succ} \\
  & \textbf{ assert } O_t = \texttt{i32} \\
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \name \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 sub "} \conc O_l \conc \texttt{", 1"} \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{pred} \\
  & \textbf{ assert } O_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

The following rule specifies code generation for the unary
minus. Since LLVM does not have a specific statement for such
operation, the translation uses subtraction:
\begin{framed}
\begin{align}
\begin{split}
 \trad{}{exp} : & \textsf{ Node} \times dict  \rightarrow text \times text \times  text\\
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \name \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 sub 0, "} \conc O_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{-} \\
  & \textbf{ assert } O_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

Also, LLVM does not have a predefined operation for the ``power''
operator provided by B0. The corresponding rule introduces a call
to a function, named \texttt{@power} that implements this operator (the LLVM source code
for such function is not presented herein):
\begin{framed}
\begin{align}
\begin{split}
 \trad{}{rel} : & \textsf{ Node} \times dict  \rightarrow text \times text \\
  \trad{n, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
  & \quad \textbf{ and } l = \name \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad l \conc \texttt{" = call i32 @power(i32, "} \conc O_l \conc \texttt{", i32 "} \conc O'_l \conc \texttt{")"}) \conc \nl), \\
  & \quad l, \texttt{"i32"} \\
  & \textbf{ where } \textsf{r.k} = \textsf{Power} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

@@

\begin{framed}
\begin{align}
\begin{split}
 \trad{}{rel} : & \textsf{ Node} \times dict  \rightarrow text \times text \\
  \trad{r, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{r.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{r.right, $D$}{exp} \\
  & \quad \textbf{ and } l = \name \textbf{ and } op = D_{op}(\textsf{r.op}) \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad l \conc \texttt{" = icmp "} \conc op \conc \texttt{" "} \conc O_t \conc \texttt{" "} \conc O_l \conc^, O'_l \conc \nl), \\
  & \quad l \\
  & \textbf{ where } \textsf{r.k} = \textsf{BinRel} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}
